````markdown
---
title: "Getting Started: Initializing an AquaCrop Project"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Getting Started: Initializing an AquaCrop Project}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

`aquacropr` can initialize a complete AquaCrop project in one call.

`init_aquacrop()` creates the expected project structure, installs the
AquaCrop executable, and prepares the workspace for the rest of the
package workflow.

It also manages the AquaCrop version used by the project. You can install
the latest official release, pin a specific release for reproducibility,
or build the current development version from source.

Downloaded binaries are cached locally and can be reused across
projects.

A minimal setup is:

```{r quick-start}
library(aquacropr)

proj_dir <- init_aquacrop(
  path = "~/my_project",
  use_rproject = FALSE
)

setwd(proj_dir)
```

`init_aquacrop()` returns the normalized project path invisibly. Keeping
this value is useful because many `aquacropr` functions use paths relative
to the project root.

The project is now ready for writing AquaCrop inputs and running
simulations with `run_aquacrop()`.

## Quick start

The usual workflow has three steps.

### 1. Load `aquacropr`

```{r library}
library(aquacropr)
```

### 2. Initialize the project

```{r init}
proj_dir <- init_aquacrop(
  path = "~/my_project",
  use_rproject = FALSE
)
```

With the default `version = NULL`, `init_aquacrop()` installs the latest
supported official AquaCrop release for the selected platform.

When a prebuilt binary is available, no compiler is required.

The function also creates the directories used for climate, crop, soil,
management, calendar, project, and output files.

### 3. Work from the project root

Most writer and reader functions use project-relative paths by default.

For example, climate files are usually written below `CLIMATE/`, while
project master files are stored below `LIST/`.

The simplest workflow is therefore:

```{r setwd}
setwd(proj_dir)
```

From this point onward, package defaults can be used directly:

```{r run}
# Write AquaCrop input files.
# ...

run_aquacrop()
```

If changing the working directory is not appropriate, such as in a
script managing several projects, pass `base_path = proj_dir` to
functions that support it.

## Choosing an AquaCrop version

The `version` argument controls which AquaCrop executable is installed.

This allows exploratory projects to follow the latest release while
reproducible analyses remain tied to a specific AquaCrop version.

### Latest release

The default is:

```{r version-null}
init_aquacrop(
  path = "~/my_project",
  version = NULL
)
```

`version = NULL` resolves to the latest supported official AquaCrop
release for the selected platform.

This is the most convenient option for new projects.

### Pinned release

For reproducible work, specify the version explicitly:

```{r version-pin}
init_aquacrop(
  path = "~/my_project",
  version = "7.1"
)
```

Pinning the version makes the AquaCrop executable part of the recorded
computational environment.

This is useful for published studies, validation exercises, operational
workflows, and any analysis that must be reproduced later.

### Development version

Use `version = "dev"` to build AquaCrop from the current development
source:

```{r version-dev}
init_aquacrop(
  path = "~/my_project",
  version = "dev"
)
```

This is intended for testing features that are not yet available in an
official release.

Unlike binary installation, the development version requires a working
Fortran toolchain.

Internally, `init_aquacrop()` delegates installation to
`install_binaries()`, which uses `install_source()` when
`version = "dev"`.

`init_aquacrop()` only exposes the arguments needed for normal project
initialization. More specialized build options can be passed directly to
`install_binaries()` or `install_source()`.

For example:

```{r version-dev-direct}
install_binaries(
  version = "dev",
  path = "~/my_project",
  compiler = "gfortran",
  keep_source = TRUE
)
```

See [Installing AquaCrop from source](#installing-aquacrop-from-source)
for the required build tools.

## Binary caching

Official AquaCrop binaries are stored in the package cache:

```r
rappdirs::user_cache_dir("aquacropr")
```

If the same version is requested again, `aquacropr` reuses the cached
archive instead of downloading it again.

```{r cache}
init_aquacrop(
  path = "~/project_a",
  version = "7.1"
)

init_aquacrop(
  path = "~/project_b",
  version = "7.1"
)
```

The first call downloads the binary when needed. The second project can
reuse the local copy.

Versions are cached independently, so several AquaCrop releases can
coexist on the same machine.

Use `force = TRUE` to bypass the cached installation and request a fresh
one:

```{r force-download}
init_aquacrop(
  path = "~/my_project",
  version = "7.1",
  force = TRUE
)
```

This is useful when repairing an installation or refreshing the
executable.

## Reinitializing a project

If `path` already contains a complete AquaCrop project,
`init_aquacrop()` leaves the existing input files unchanged.

It checks the project structure and installs the executable only when it
is missing.

Two arguments allow this behavior to be overridden.

`force = TRUE` reinstalls the AquaCrop executable.

`overwrite = TRUE` reruns project initialization. Missing directories are
recreated, bundled simulation templates are restored to `SIMUL/`, and
`README.txt` is rewritten.

Existing user input files such as `.PLU`, `.CRO`, `.SOL`, and `.PRM`
files are not deleted.

```{r overwrite}
init_aquacrop(
  path = "~/my_project",
  force = TRUE,
  overwrite = TRUE
)
```

This is useful after an incomplete initialization or when the project
needs to be prepared again for another platform.

## RStudio project behavior

By default, `use_rproject = TRUE`.

When `init_aquacrop()` is called from an active interactive RStudio
session, it can create and open an `.Rproj` file through
`rstudioapi::openProject()`.

Opening the project changes the active RStudio session. This is useful
for interactive work, but less suitable inside scripts that are expected
to continue running after initialization.

For scripted workflows, use:

```{r rproject}
proj_dir <- init_aquacrop(
  path = "~/my_project",
  use_rproject = FALSE
)

setwd(proj_dir)
```

This is the recommended pattern for batch scripts, CI pipelines,
containers, servers, and other non-interactive workflows.

Outside an active RStudio session, the RStudio-specific branch is
skipped. No `.Rproj` file is created.

The AquaCrop project itself does not depend on RStudio.

## Selecting the target platform

The operating system is detected automatically.

```{r os-auto}
init_aquacrop(path = "~/my_project")
```

In most cases, no `os` argument is needed.

The target platform can also be selected explicitly from `"windows"`,
`"linux"`, or `"macos"`.

```{r os}
init_aquacrop(
  path = "~/my_project",
  os = "windows"
)
```

This is useful when preparing a project on one operating system for
execution on another.

For example, a project can be organized on Linux or macOS while
installing the Windows AquaCrop executable that will later be used on a
Windows machine.

The project directory structure remains the same across platforms.

## Installing AquaCrop from source {#installing-aquacrop-from-source}

When `version = "dev"` is requested, `install_binaries()` delegates the
build to `install_source()`.

The source installation performs the main steps automatically:

1. detect the Fortran compiler;
2. download the AquaCrop source;
3. compile the executable;
4. install it in the requested directory; and
5. remove temporary build files.

Compiler detection uses:

```text
R CMD config FC
```

This retrieves the Fortran compiler configured for the current R
installation.

A compiler can also be supplied directly when calling
`install_binaries()` or `install_source()`.

Temporary source files are removed after a successful build unless
`keep_source = TRUE`.

### Linux

A working `make` installation and a compatible Fortran compiler are
required.

For example:

```sh
sudo apt-get install make gfortran
```

or:

```sh
sudo dnf install make gcc-gfortran
```

### macOS

A Fortran compiler must be installed separately.

The `clang` compiler provided with Xcode does not include Fortran
support.

For R-based workflows, a compatible `gfortran` installation should be
used.

### Windows

A compatible Rtools installation provides the build tools required for
source compilation, including `make` and `gfortran`.

Compiler requirements can change with the development source. Users
building `version = "dev"` should therefore ensure that their local
toolchain is compatible with the current AquaCrop source.

## Project structure

A successful call to `init_aquacrop()` creates the following workspace:

```text
my_project/
├── my_project.Rproj
├── aquacrop[.exe]
├── README.txt
├── CAL/
├── CLIMATE/
├── CROP/
├── LIST/
├── MANAGEMENT/
├── OBS/
├── OUTP/
├── PARAM/
├── SIMUL/
└── SOIL/
```

The `.Rproj` file is only created when the RStudio integration is used.

The directories group AquaCrop files by role.

`CAL/` contains growing-season calendar files.

`CLIMATE/` contains rainfall, temperature, reference
evapotranspiration, climate descriptor, and atmospheric carbon dioxide
files.

`CROP/` contains crop parameter files.

`SOIL/` contains soil profiles, initial soil-water conditions, and
groundwater files.

`MANAGEMENT/` contains field-management and irrigation inputs.

`LIST/` contains project master files.

`OBS/` contains observed data used for comparison or model evaluation.

`OUTP/` contains simulation outputs.

`PARAM/` contains AquaCrop program parameters.

`SIMUL/` contains simulation settings.

The AquaCrop executable is installed at the project root.

## Why `aquacropr` uses this structure

The project layout follows AquaCrop conventions while making input types
explicit.

The official standalone traditionally groups many inputs in a common
`DATA/` directory. This works well for interactive use, but becomes less
convenient when projects contain many sites, crops, climate datasets, or
simulation configurations.

`aquacropr` separates these inputs into dedicated directories:

```text
CAL/
CLIMATE/
CROP/
MANAGEMENT/
SOIL/
```

The AquaCrop file formats themselves are unchanged.

The difference is organizational.

Predictable paths make files easier to generate, inspect, automate, and
version-control from R. They also match the relative-path conventions
used by the package.

Core AquaCrop directories such as `LIST/`, `PARAM/`, `SIMUL/`, and
`OUTP/` remain part of the project.

The result is an AquaCrop-compatible workspace organized for scripted
and reproducible workflows.

## Reproducible project setup

For exploratory work, the latest supported release is convenient:

```{r reproducible-latest}
proj_dir <- init_aquacrop(
  path = "~/my_project",
  use_rproject = FALSE
)
```

For reproducible analyses, pin the AquaCrop version:

```{r reproducible-pinned}
proj_dir <- init_aquacrop(
  path = "~/my_project",
  version = "7.1",
  use_rproject = FALSE
)
```

This makes the executable version explicit alongside the R package,
scripts, and input data used to construct the simulation.

`aquacropr` does not replace AquaCrop. It manages the project,
executable, files, and simulation workflow around it.

Once initialized, the same workspace can be populated programmatically
with climate, crop, soil, management, and calendar inputs and then passed
to `run_aquacrop()`.

## Next steps

Project initialization normally needs to be performed only once.

The usual workflow is then to:

1. work from the project root or pass `base_path` explicitly;
2. prepare the required AquaCrop input files;
3. write the project master file;
4. run AquaCrop; and
5. read and analyse the simulation outputs.

The rest of the `aquacropr` workflow builds on the project structure
created by `init_aquacrop()`.
````
