---
title: "Sowing Dates: Extensible Onset Detection in aquacropr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sowing Dates: Extensible Onset Detection in aquacropr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval    = FALSE,
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

## Overview

AquaCrop requires a planting date for each simulation year. For rainfed
crops, this date is often derived from the onset of the rainy season.

The AquaCrop standalone provides a fixed set of criteria for determining
this onset. These criteria cover common rainfall- and temperature-based
rules, but they cannot represent every agronomic definition of the
beginning of a growing season.

This becomes important in rainfall-limited environments. In the Sahel,
for example, the first substantial rainfall event may be followed by a
long dry spell. A rainfall threshold can therefore identify a date that
is meteorologically plausible but unsuitable for crop establishment.

`aquacropr` addresses both issues by moving onset detection into R.

The package reproduces the native AquaCrop criteria, but is not limited
to them. The same framework can implement onset definitions based on
rainfall accumulation, dry spells, temperature, multiple conditions, or
other indicators derived from daily climate data.

Three additional methods are currently implemented for Sahelian
applications: Sivakumar (1988), Marteau et al. (2009), and a fuzzy
criterion based on Waongo et al. (2014). They illustrate the main added
value of the approach: **the onset criterion becomes extensible, while
the resulting simulation remains compatible with AquaCrop**.

This vignette describes the onset engine, the available criteria, and
the principles for extending it.

> **Reproducibility**
>
> Code in this vignette is shown but not executed. The complete dataset
> used to reproduce the examples is distributed separately because the
> archive is approximately 1 GB.
>
> **Data:** hosted on [OSF](https://osf.io) as `TEST.zip`
> (`<OSF_DIRECT_DOWNLOAD_URL>` -- see the "Regional Crop Simulations"
> vignette's "Download the example dataset" section for the download
> code and archive contents).

## From an AquaCrop criterion to an extensible onset engine

### The AquaCrop workflow

The `.CAL` file describes how the beginning of the growing season should
be determined.

In the AquaCrop GUI, this information is used to derive sowing dates
before the project master file (`.PRM`) is written. For the standalone
simulation, the essential result is the planting date stored in the
`.PRM`.

This creates a useful separation between two tasks:

1. determining *when planting should occur*; and
2. simulating the crop from that planting date.

`aquacropr` uses this separation to make the first task extensible.

### The `aquacropr` approach

`find_onset()` implements onset detection in R.

It reads the `.CAL` configuration and the required climate data, applies
the selected criterion to each year, and returns the corresponding
planting dates.

`write_prm()` uses the same engine to place these dates in the `.PRM`
file expected by AquaCrop.

The simulation itself does not need to know how the planting date was
obtained.

This is the key design choice.

A planting date can come from a native AquaCrop criterion, a published
agroclimatic definition, or a more specialized rule implemented in
`aquacropr`. Once the date has been determined, the downstream AquaCrop
workflow is unchanged.

The package therefore preserves compatibility with AquaCrop while
removing the onset calculation from a fixed set of criteria.

```text
Climate data
     |
     v
   .CAL
     |
     v
aquacropr onset engine
     |
     +-- Native AquaCrop criteria
     |
     +-- Sivakumar
     |
     +-- Marteau
     |
     +-- Fuzzy criterion
     |
     +-- Future criteria
     |
     v
Planting date for each year
     |
     v
   .PRM
     |
     v
AquaCrop standalone
```

This architecture also makes onset detection easier to inspect and test.
The dates can be examined before a crop simulation is launched.

```{r workflow}
aquacropr::write_cal(
  site_name = "Ouaga",
  onset = "rainfall",
  window_start = 100,
  window_length = 120,
  criterion = 2,
  preset_value = 20,
  successive_days = 3
)

aquacropr::find_onset("Ouaga")

aquacropr::write_prm(
  site_name = "Ouaga",
  ...
)
```

Calling `find_onset()` explicitly is optional. It is useful when onset
dates need to be inspected, compared, or validated before running the
simulation.

## Available onset methods

The current implementation supports fixed, rainfall, and thermal onset.

| `onset` | Climate file | Typical use |
|---|---|---|
| `"fixed"` | `.PLU` | Prescribed planting date |
| `"rainfall"` | `.PLU` | Rainfed crops |
| `"thermal"` | `.Tnx` | Temperature-limited seasons |

Within these types, `aquacropr` reproduces the native AquaCrop criteria
and adds methods that are not available in the standalone.

| Method | Main information used | AquaCrop standalone |
|---|---|---|
| Native criteria 1--4 | Rainfall or temperature | Yes |
| Sivakumar (1988) | Rainfall + subsequent dry spell | No |
| Marteau et al. (2009) | Planting rainfall + subsequent dry spell | No |
| Fuzzy criterion | Rainfall + wet days + dry-spell risk | No |

The distinction between **native** and **additional** criteria is
important, but the framework is not restricted to the methods in this
table.

A new onset definition can combine any information available to the
onset engine. For example, a criterion could incorporate several
rainfall thresholds, antecedent conditions, dry-spell characteristics,
temperature, evapotranspiration, or other climate-derived indicators.

The current Sahelian methods should therefore be viewed both as useful
criteria in their own right and as examples of what becomes possible
once onset detection is handled in R.

## Why additional criteria matter

An onset criterion is an operational definition of a planting
opportunity.

Different definitions answer different questions.

A cumulative rainfall criterion asks whether enough rain has fallen.

A multi-day criterion asks whether enough rain has fallen within a short
period.

A rainfall-to-ETo criterion also considers atmospheric demand.

A false-start criterion asks an additional question: whether favourable
conditions persist after the candidate planting date.

A fuzzy criterion can go further by allowing several indicators to
contribute gradually rather than making the decision depend on a single
hard threshold.

These differences matter in the Sahel because early-season rainfall is
highly variable. A substantial rainfall event can trigger germination
and still be followed by conditions unable to sustain crop
establishment.

The additional methods in `aquacropr` demonstrate how the same AquaCrop
simulation can be driven by onset definitions adapted to this type of
agroclimatic problem.

## Extending the onset framework

The three additional methods are not intended to define the limit of the
onset engine.

They demonstrate three forms of extension:

- **Sivakumar** adds a condition after the initial rainfall trigger;
- **Marteau** represents an onset definition informed by observed
  planting behaviour;
- **the fuzzy criterion** combines several indicators through continuous
  membership functions rather than hard thresholds.

Other definitions can follow the same principle.

For example, future criteria could use soil-water conditions, different
dry-spell definitions, crop-specific establishment requirements, or
locally calibrated combinations of climate indicators.

The important point is that these extensions affect the calculation of
the planting date, not the AquaCrop simulation interface.

As long as the onset engine produces a valid planting date for each
simulation year, `write_prm()` can pass that result to AquaCrop in the
same way.

This makes onset detection a modular part of the `aquacropr` workflow
rather than a closed list of planting rules.
